A generalization of the Bradley-Terry model for draws in chess

### 2019-05-17 inst/collusion_chess_draws.Rnw
work

In inference problems where the dataset is outcomes of paired
comparisons, such as chess, the Bradley-Terry model can be augmented
to chess-specific considerations in a number of ways.  Here I present
a new generalization of Bradley-Terry in which a chess game is
regarded as a three-way competition between the two players and an
entity that wins if the game is drawn.  Bradley-Terry can be further
modified to account for the white player advantage in a number of
ways.  Such reification affords insight into players' strengths,
response to playing black or white, and risk-aversion as manifested by
probability of drawing.  The likelihood functions arising are easily
optimised numerically.  I analyse a number of datasets of chess
results, including the infamous World Chess Championships 1963 in
which the Soviets were accused of collusion.  I conclude that the
scoreline of the Candidates Tournament (Cura\c{c}ao 1962) exhibits no
evidence of collusion, in agreement with previous work.  However, I
find strong evidence of increased Soviet proclivity to draw in the
Interzonal Tournament, Stockholm 1962.
}

@@

\paragraph{Draws} Chess games may be won by either player, or may be
drawn; the implication of a draw is that neither player is better than
the other.  Before 1867, tournament games that were drawn were
replayed and in this context we note that this procedure reduces a
chess game to a Bernoulli trial, with probability one.  This system
became unworkable due to the large number of draws, resulting in the
modern practice of awarding each player of a drawn game half a point,
rendering chess a zero-sum game.  Drawn games are generally considered
to be undesirable as they are usually poorer in spectacular or
striking elements such as sharp tactical combinations or sacrifice.
Excessive drawing dissuades sponsors, leaves spectators feeling
cheated, and gives commentators less material to discuss.  Players
themselves do not seem to value draws in the wider context of the
game: \citeauthor{fischer1969}, for example, includes only 8 draws in
his compilation of 60 memorable games, surely not representative of
tournament play in which about 39.4\% of games are drawn (binomial
test, $p<10^{-5}$).

@@

The above likelihood functions are objective and reasonable
representations of chess results.  However, they differ markedly from
previous likelihood functions of draws in that they include entities
representing proclivity to draw $D$, and first-mover advantage ($W$).

@@

\begin{equation}
\parbox{4in}{$
P^5\,
{Ke}^5\,
G^4
\mathit{Fis}\,  % Fischer, cf Filip
\mathit{Ko}^5
B
T^2  % Tal
\mathit{Fil}\,
D^{42}C^{12}\oo\\ \rule{10mm}{0mm}
\left(P+W\right)^3
\left(\mathit{Ke}+W\right)^4
\left(G+W\right)^4\
\left(\mathit{Fis}+W\right)^7\oo\\ \rule{20mm}{0mm}
\left(\mathit{Ko}+W\right)^2
\left(B+W\right)^5
\left(T+W\right)
\left(\mathit{Fil}+W\right)\oo
$}
\end{equation}

on the numerator (where $C$ is the strength of a collusive draw), and
a total of 36 terms such as $\left(P + \mathit{Fis}+W+D\right)^{4}$
and $\left(P + \mathit{Ke}+W+C\right)^{4}$ on the denominator.

@@

\begin{figure}[htbp]
\begin{center}
<<collusion,echo=TRUE,fig=TRUE>>=
dotchart(maxp(collusion),pch=16)
@
\caption{sds\label{collusion_fig}}
  \end{center}
\end{figure}

In figure~\ref{collusion_fig}, note the high estimate for $D_s$ (also
note
Fischer's dominance).

@@

<<echo=FALSE,print=FALSE>>=
do_from_scratch  <- FALSE  # set to TRUE to recalculate (slow, maybe 2 mins)
hotstart <- indep(maxp(collusion))

@@

\section{Conclusions}

@@

>>>>

In inference problems where the dataset is outcomes of paired
comparisons, such as chess, the Bradley-Terry model can be augmented
to chess-specific considerations in a number of ways.  Here I present
a new generalization of Bradley-Terry in which a chess game is
regarded as a three-way competition between the two players and an
entity that wins if the game is drawn.  Bradley-Terry can be further
modified to account for the white player advantage in a number of
ways.  Such reification affords insight into players' strengths,
response to playing black or white, and risk-aversion as manifested by
probability of drawing.  The likelihood functions arising are easily
optimised numerically.  I analyse a number of datasets of chess
results, including the infamous World Chess Championships 1963 in
which the Soviets were accused of collusion.  I conclude, on the basis
of a dataset that includes only the scorelines at the event itself,
that the scoreline of the Candidates Tournament (Cura\c{c}ao 1962)
exhibits evidence of collusion ($p<10^{-5}$), in agreement with
previous work that uses additional information such as historical
results including number of moves.}

@@

\paragraph{Draws} Chess games may be won by either player, or may be
drawn; the implication of a draw is that neither player is better than
the other.  Before 1867, tournament games that were drawn were
replayed until one or other player won, thus reducing chess game to a
Bernoulli trial, almost surely.  This system became unworkable due to
the large number of draws, resulting in the modern practice of
awarding each player of a drawn game half a point, rendering chess a
zero-sum game.  Drawn games are generally considered to be undesirable
as they are usually poorer in spectacular or striking elements such as
sharp tactical combinations or sacrifice.  Excessive drawing dissuades
sponsors, leaves spectators feeling cheated, and gives commentators
less material to discuss.  Players themselves do not seem to value
draws in the wider context of the game: \citeauthor{fischer1969}, for
example, includes only 8 draws in his compilation of 60 memorable
games, surely not representative of tournament play in which about
39.4\% of games are drawn (binomial test, $p<10^{-5}$).

@@

The above likelihood functions are objective and reasonable
representations of chess results.  However, they differ markedly from
previous likelihood functions of draws in that they include reified
entities representing proclivity to draw $D$, and white or first-mover
advantage ($W$).

@@

\begin{equation}\parbox{4in}{$
P^5\,
{Ke}^5\,
G^4
\mathit{Fis}\,  % Fischer, cf Filip
\mathit{Ko}^5
B
T^2  % Tal
\mathit{Fil}\,
D^{42}C^{12}\oo\\ \rule{10mm}{0mm}
\left(P+W\right)^3
\left(\mathit{Ke}+W\right)^4
\left(G+W\right)^4\
\left(\mathit{Fis}+W\right)^7\oo\\ \rule{20mm}{0mm}
\left(\mathit{Ko}+W\right)^2
\left(B+W\right)^5
\left(T+W\right)
\left(\mathit{Fil}+W\right)\oo
$}\end{equation}

\noindent on the numerator (where $C$ is the strength of a collusive
draw), and a total of 36 terms such as
$\left(P+\mathit{Fis}+W+D\right)^{4}$ and
$\left(P+\mathit{Ke}+W+C\right)^{4}$ on the denominator.  The
likelihood ratio for the hypothesis that collusive draws have the same
strength as regular draws isabout $e^{11.8}\simeq 10^5$: we conclude
that the high number of draws among the three accused players was
unlikely to be due to chance.

One plausible explanation for the previ

@@

@@

<<echo=FALSE,print=FALSE>>=
do_from_scratch  <- FALSE  # set to TRUE to recalculate (slow, maybe 2 mins)
hotstart <- indep(maxp(interzonal_collusion))

@@

\section{Conclusions and further work}

The present work details an approach to chess analysis based on
reification of entities representing propensity to draw, and the
first-mover (or white player) advantage.  Likelihood functions for the
players' and reified entities' strengths are defined using only
specific tournament outcomes; further, the analysis uses only game
outcome, not length.

@@

\cite{moul2009} present detailed Monte Carlo simulations to assess the likely impact of any Soviet cartel on tournament outcomes.  In principle, the present system of likelihood and reified entities would admit such simulations and might be interesting to compare results.
